### 2019-07-05 methodology/bayesian-networks.tex
Sezione sull'algoritmo Pseudo-MPE

\subsection{Information Theory}
The birth of the field of \textit{information theory} is usually traced back to the seminal paper \enquote{A Mathematical Theory of Communication} (\cite{Shannon1949}) where Claude Shannon set the mathematical basis for the quantification of the amount of \textit{information} transmissible over a noisy channel. 
In his words \enquote{The fundamental problem of communication is that of reproducing at one point, either exactly or approximately, a message selected at another point.}
The concepts of field are broad enough to have influenced practically every other scientific discipline and deep enough to have enabled the \enquote{digital age}, for example by enabling the creation of ever more complicated coding schemes for the compression, reconstruction and obfuscation of digital data.

@@

\subsection{Graph Theory}
Many problems in Machine Learning (ML) don't involve classification or prediction of single data points in isolation, but of set of entities that may present a more, or less, complex relation with each other. 
Most real-world phenomena fit into the latter framework.
Graphs are one of the most powerful tools for the modelling of this class of problems, as their structure naturally captures the wide variety of relations that may exist between entities.
These range from the atomical structure of a molecule to a social network of friends.  
In both these examples graphs help in reasoning, visualising and making inferences and predictions.

@@

>>>>

\subsection{Information Theory} \label{subsec:information-theory}
The birth of the field of \textit{information theory} is usually traced back to the seminal paper \enquote{A Mathematical Theory of Communication} (\cite{Shannon1949}) where Claude Shannon set the mathematical basis for the quantification of the amount of \textit{information} transmissible over a noisy channel. 
In his words \enquote{The fundamental problem of communication is that of reproducing at one point, either exactly or approximately, a message selected at another point.}
The concepts of field are broad enough to have influenced practically every other scientific discipline and deep enough to have enabled the \enquote{digital age}, for example by enabling the creation of ever more complicated coding schemes for the compression, reconstruction and obfuscation of digital data.

@@

\subsection{Graph Theory} \label{subsec:graph-theory}
Many problems in Machine Learning (ML) don't involve classification or prediction of single data points in isolation, but of set of entities that may present a more, or less, complex relation with each other. 
Most real-world phenomena fit into the latter framework.
Graphs are one of the most powerful tools for the modelling of this class of problems, as their structure naturally captures the wide variety of relations that may exist between entities.
These range from the atomical structure of a molecule to a social network of friends.  
In both these examples graphs help in reasoning, visualising and making inferences and predictions.

@@

\textbf{A very important thing to note is that the greedy assignment where each variable picks its most likely value can be very different from the most likely joint assignment of all variables.}

### 2019-07-05 methodology/data-set.tex
Sezione sull'algoritmo Pseudo-MPE

\subsection{Istituto Cantonale di Patologia}
Istituto Cantonale di Patologia (ICP) is an institute based in Locarno that is specialised in the histological analysis of tissue samples received from private patients, clinics and hospitals, mainly in support of cancer diagnosis.
These tests are aimed at identifying the precise profile of the cancer cells and thus inform the clinician on the best treatment for the specific patient.

@@

\todo{chiedere a Vittoria martedì cosa si aspettavano/aspettano}

My interest in collaborating with the Istituto stemmed from the desire to apply the methods described in Sec. \ref{sec:novel-contributions} to a real-world case.
Being the theoretical work being carried out in this thesis an expert-driven MPE approximation, collaboration with the institute has also provided the opportunity to implement a proof of concept using real histological data.
The doctors and researchers of the Istituto have been able to validate the model software that I have developed from an Explainable AI point of view.
That is to say, they have validated the capacity of the developed software to support clinical decision making and surface clarifying explanations of the data set.

\todo{da confermare con Vittoria più avanti}

\subsection{Provided Data Set}
The data set I was provided with consists of the histological records of 3218 breast cancer patients.
The data set had been pre-processed by collaborators of the ICT with some of the values being compiled by hand. 
In Tab. \ref{tab:datasetvariables} is a description of the measured variables, together with their clinical meaning.
The value distribution of the data set is shown in Tab. \ref{tab:datasetdistribution}.

@@

\begin{table*}[htbp]
\caption{Data set variables}
\begin{tabularx}{\textwidth}{@{} l Y @{}}
\toprule 
\textbf{Variable} & Meaning \\
\midrule 
\textbf{Codice globale} & Unique patient identifier \\
\textbf{mut17q21} & \\
\textbf{loss 17} & \\
\textbf{et\`a arrotondata} & The age of the patient at diagnosis\\
\textbf{Lateralit\`a} & The affected breast \\
\textbf{Situ SUBGROUP MZ} & The site code of the tumour \\
\textbf{Morfologia SUBGROUP MZ} & The morphology classification of the tumour \\
\textbf{pT SUBGROUP MZ} & Primary tumour in the TNM classification for breast cancer \\
\textbf{pN SUBGROUP MZ} & Pathologic in the TNM classification for breast cancer \\
\textbf{M 8.2.96} & Distant metastasis in the TNM classification for breast cancer \\
\textbf{Differenziazione} &  \\
\textbf{Recettori estrogeni percento 1.1.2003} & \\
\textbf{Recettori progestinici percento 1.1.2003} & \\
\textbf{c erbB 2  cod percento 1.1.2003} & \\
\textbf{Ki67 cod percento} & \\
\textbf{FISHRatio} & \\
\bottomrule
\end{tabularx}
\label{tab:datasetvariables}
\end{table*}

\begin{table*}[htbp]
\caption{Data set distribution}
\begin{tabularx}{\textwidth}{@{} l X c @{}}
\toprule 
\textbf{Variable} & Unique values & Distribution \\
\midrule 
\textbf{mut17q21} & 2 & \includegraphics[width=0.2\textwidth, height=10mm]{methodology/images/mut17q21}  \\
\textbf{loss 17} & 3 & \includegraphics[width=0.2\textwidth, height=10mm]{methodology/images/loss_17}\\
\textbf{eta arrotondata} & 74 & \includegraphics[width=0.2\textwidth, height=10mm]{methodology/images/eta_arrotondata}\\
\textbf{ateralit\`a} & 3 & \includegraphics[width=0.2\textwidth, height=10mm]{methodology/images/lateralita} \\
\textbf{situ} & 5 & \includegraphics[width=0.2\textwidth, height=10mm]{methodology/images/situ} \\
\textbf{morfologia} & 5 & \includegraphics[width=0.2\textwidth, height=10mm]{methodology/images/morfologia} \\
\textbf{pT} & 23 & \includegraphics[width=0.2\textwidth, height=10mm]{methodology/images/pt} \\
\textbf{pN} & 6 & \includegraphics[width=0.2\textwidth, height=10mm]{methodology/images/pn} \\
\textbf{M} & 3 & \includegraphics[width=0.2\textwidth, height=10mm]{methodology/images/m} \\
\textbf{differenziazione} & 5 & \includegraphics[width=0.2\textwidth, height=10mm]{methodology/images/differenziazione}  \\
\textbf{recettori estrogeni} & 40 & \includegraphics[width=0.2\textwidth, height=10mm]{methodology/images/recettori_estrogeni} \\
\textbf{recettori progestinici} & 40 & \includegraphics[width=0.2\textwidth, height=10mm]{methodology/images/recettori_progestinici}\\
\textbf{c erbB 2} & 4 & \includegraphics[width=0.2\textwidth, height=10mm]{methodology/images/c_erb_2}\\
\textbf{Ki67} & 52 & \includegraphics[width=0.2\textwidth, height=10mm]{methodology/images/ki67}\\
\textbf{FISH} & 5 & \includegraphics[width=0.2\textwidth, height=10mm]{methodology/images/fish}\\
\bottomrule
\end{tabularx}
\label{tab:datasetdistribution}
\end{table*}

\begin{table*}[htbp]
\caption{Data set preprocessing}
\begin{tabularx}{\textwidth}{@{} l Y @{}}
\toprule 
\textbf{Variable} & Action \\
\midrule 
\textbf{Codice globale} & Remove variable \\
\textbf{mut17q21} & Remove blanks \\
\textbf{loss 17} & Remove blanks \\
\textbf{eta arrotondata} & Bin into \enquote{$< 40$}, \enquote{$40-50$}, \enquote{$\geq 50$}\\
\textbf{lateralita} & Remove blanks and \enquote{sconosciuta} \\
\textbf{situ} & Remove blanks \\ \addlinespace
\textbf{morfologia} & Remove blanks and \enquote{unuseful} if performance on classification is subpar \\ \addlinespace
\textbf{pT} & Remove blanks and \enquote{unuseful}  \\
\textbf{pN} & Remove blanks and bin into \enquote{0} and \enquote{$\neq0$}\\
\textbf{M} & Remove blanks \\ 
\textbf{differenziazione} & Remove blanks and \enquote{Sconosciuto o non applicabile} \\ \addlinespace
\textbf{recettori estrogeni} & Remove blanks and bin into \enquote{negativo} if $\leq 10$,
		\enquote{debolmente positivo} if $\leq 50$, 
		\enquote{fortemente positivo} if $> 50$ \\ \addlinespace
\textbf{recettori progestinici} & Remove blanks and bin into \enquote{negativo} if $\leq 10$, 
		\enquote{debolmente positivo} if $\leq 50$, 
		\enquote{fortemente positivo} if $> 50$ \\ \addlinespace
\textbf{c erbB 2} & Remove blanks \\ 
\textbf{ki67} & Remove blanks and bin into \enquote{<14}, 
		\enquote{14-20}, \enquote{20-30}, \enquote{>30} \\ 
\textbf{FISH} & Remove blanks \\
\bottomrule
\end{tabularx}
\label{tab:datasetpreprocess}
\end{table*}

\section{Validation Methodology}
\todo{discutere con Vittoria cosa vogliono esattamente e come validare }

>>>>

\subsection{Istituto Cantonale di Patologia}
Istituto Cantonale di Patologia (ICP) is an institute based in Locarno that is specialised in the histological analysis of tissue samples received from private patients, clinics and hospitals, mainly in support of cancer diagnosis.
Its Laboratory of Medical Diagnostics supports pathologists in the diagnosis of neoplastic diseases through the application of cytogenetic techniques; that is, the focus is on understanding how chromosomes relate to cell behaviour, particularly during mitosis and meiosis.
One of the techniques used is Fluorescence in Situ Hybridization (FISH) that is able to localise the presence or absence of specific DNA sequences in chromosomes.
These tests are aimed at identifying the precise profile of the cancer cells and thus inform the clinician on the best treatment for the specific patient.

@@

\todo[inline]{chiedere a Vittoria cosa si aspettavano/aspettano}

My interest in collaborating with the Istituto stemmed from the desire to apply the methods described in Sec. \ref{sec:novel-contributions} to a real-world case.
Being the theoretical work being carried out in this thesis an expert-driven MPE approximation, collaboration with the institute has also provided the opportunity to implement a proof of concept using real histological data.
The doctors and researchers of the Istituto have been able to validate, from an Explainable AI and clinical relevance points of view, the model software that I have developed.
That is to say, they have validated the capacity of the developed software both in its capacity to support clinical decision making and to surface clarifying explanations of the data set and in its adherence to established medical literature.

\todo[inline]{da confermare con Vittoria piu avanti}

\subsection{Provided Data Set}
The data set I was provided with was created by \textit{Registro Tumori Ticino} (Locarno, Ticino) in order to highlight possible new relations between clinical, histopathological and molecular features, as well as to potentially discover novel biomarkers for the progression of the disease 
It consists of the histological records over 38 recorded variables of 3218 breast cancer patients who have been diagnosed between the years 2005 and 2014 within the Ticino canton of Switzerland.
The data set had been pre-processed by collaborators of IDSIA in agreement with the ICT with 13 of the variables being dropped, because not considered relevant.
In particular, all variables relating to the patient post-treatment were discarded as well as those recording the diagnosis date.
In Tab. \ref{tab:datasetvariables} is a description of the measured variables, together with their clinical meaning.
The value distribution of the data set is shown in Tab. \ref{tab:datasetdistribution}.

@@

\begin{table*}[htbp]
\caption{Data set variables}
\begin{tabularx}{\textwidth}{@{} l Y @{}}
\toprule 
\textbf{Variable} & Clinical meaning \\
\midrule 
\textbf{Codice globale} & Unique patient identifier \\
\textbf{mut17q21} & Mutation of chromosome 17 \\
\textbf{loss 17} & Loss of chromosome 17 \\
\textbf{et\`a arrotondata} & The age of the patient at diagnosis \\
\textbf{Lateralit\`a} & The affected breast \\
\textbf{Situ SUBGROUP MZ} & The primary site code of the tumour \\
\textbf{Morfologia SUBGROUP MZ} & The morphology classification of the tumour \\
\textbf{pT SUBGROUP MZ} & Primary tumour in the TNM classification for breast cancer \\
\textbf{pN SUBGROUP MZ} & Pathologic in the TNM classification for breast cancer \\
\textbf{M 8.2.96} & Distant metastasis in the TNM classification for breast cancer \\
\textbf{Differenziazione} & Tumour grade \\
\textbf{Recettori estrogeni percento 1.1.2003} & Expression of estrogen receptors \\
\textbf{Recettori progestinici percento 1.1.2003} & Expression of progestin receptors \\
\textbf{c erbB 2  cod percento 1.1.2003} & ErbB2 marker expression \\
\textbf{Ki67 cod percento} & Tumoural proliferation index \\
\textbf{FISHRatio} & FISH analysis result \\
\bottomrule
\end{tabularx}
\label{tab:datasetvariables}
\end{table*}

\begin{table*}[htbp]
\caption{Data set distribution before pre-processing}
\begin{tabularx}{\textwidth}{@{} l X c @{}}
\toprule 
\textbf{Variable} & Unique values & Distribution \\
\midrule 
\textbf{mut17q21} & 2 & \includegraphics[width=0.2\textwidth, height=10mm]{methodology/images/mut17q21}  \\
\textbf{loss 17} & 3 & \includegraphics[width=0.2\textwidth, height=10mm]{methodology/images/loss_17}\\
\textbf{eta arrotondata} & 74 & \includegraphics[width=0.2\textwidth, height=10mm]{methodology/images/eta_arrotondata}\\
\textbf{ateralit\`a} & 3 & \includegraphics[width=0.2\textwidth, height=10mm]{methodology/images/lateralita} \\
\textbf{situ} & 5 & \includegraphics[width=0.2\textwidth, height=10mm]{methodology/images/situ} \\
\textbf{morfologia} & 5 & \includegraphics[width=0.2\textwidth, height=10mm]{methodology/images/morfologia} \\
\textbf{pT} & 23 & \includegraphics[width=0.2\textwidth, height=10mm]{methodology/images/pt} \\
\textbf{pN} & 6 & \includegraphics[width=0.2\textwidth, height=10mm]{methodology/images/pn} \\
\textbf{M} & 3 & \includegraphics[width=0.2\textwidth, height=10mm]{methodology/images/m} \\
\textbf{differenziazione} & 5 & \includegraphics[width=0.2\textwidth, height=10mm]{methodology/images/differenziazione}  \\
\textbf{recettori estrogeni} & 40 & \includegraphics[width=0.2\textwidth, height=10mm]{methodology/images/recettori_estrogeni} \\
\textbf{recettori progestinici} & 40 & \includegraphics[width=0.2\textwidth, height=10mm]{methodology/images/recettori_progestinici}\\
\textbf{c erbB 2} & 4 & \includegraphics[width=0.2\textwidth, height=10mm]{methodology/images/c_erb_2}\\
\textbf{Ki67} & 52 & \includegraphics[width=0.2\textwidth, height=10mm]{methodology/images/ki67}\\
\textbf{FISH} & 5 & \includegraphics[width=0.2\textwidth, height=10mm]{methodology/images/fish}\\
\bottomrule
\end{tabularx}
\label{tab:datasetdistribution}
\end{table*}

\begin{table*}[htbp]
\caption{Data set preprocessing steps}
\begin{tabularx}{\textwidth}{@{} l Y @{}}
\toprule 
\textbf{Variable} & Action \\
\midrule 
\textbf{Codice globale} & Remove variable \\
\textbf{mut17q21} & Remove blanks \\
\textbf{loss 17} & Remove blanks \\
\textbf{eta arrotondata} & Bin into \enquote{$< 40$}, \enquote{$40-50$}, \enquote{$\geq 50$} \\
\textbf{lateralita} & Remove blanks and \enquote{sconosciuta} \\
\textbf{situ} & Remove blanks \\ \addlinespace
\textbf{morfologia} & Remove blanks and \enquote{unuseful} if performance on classification is subpar \\ \addlinespace
\textbf{pT} & Remove blanks and \enquote{unuseful}  \\
\textbf{pN} & Remove blanks and bin into \enquote{0} and \enquote{$\neq0$}\\
\textbf{M} & Remove blanks \\ 
\textbf{differenziazione} & Remove blanks and \enquote{Sconosciuto o non applicabile} \\ \addlinespace
\textbf{recettori estrogeni} & Remove blanks and bin into \enquote{negativo} if $\leq 10$,
		\enquote{debolmente positivo} if $\leq 50$, 
		\enquote{fortemente positivo} if $> 50$ \\ \addlinespace
\textbf{recettori progestinici} & Remove blanks and bin into \enquote{negativo} if $\leq 10$, 
		\enquote{debolmente positivo} if $\leq 50$, 
		\enquote{fortemente positivo} if $> 50$ \\ \addlinespace
\textbf{c erbB 2} & Remove blanks \\ 
\textbf{ki67} & Remove blanks and bin into \enquote{<14}, 
		\enquote{14-20}, \enquote{20-30}, \enquote{>30} \\ 
\textbf{FISH} & Remove blanks \\
\bottomrule
\end{tabularx}
\label{tab:datasetpreprocess}
\end{table*}

\section{Validation Methodology}
\todo[inline]{discutere con Vittoria come validare al meglio}
Having direct access to expert pathologists has not only helped in guiding research into the theoretical explainability properties of the system but also enabled their validation.
There are two main validation streams to be addressed: from the clinical point of view and from the explainability one, with the results of the latter depending on those of the former.

\subsection{Clinical validation}
A validation of the methods carried out in this thesis in their adherence to established clinical literature, is of paramount importance.
A failure on the Bayesian Network's part in capturing the true relationships between the variables would hamper it in being able to give any meaningful representation of them.
For the expert to even start to trust the system or to be able to make sense of its outputs, it is vital that there be as little cognitive dissonance between the experts' basic beliefs and expectations and those that he sees represented in the system.

For this reason, the initial validation phase with the Istituto Cantonale di Patologia concentrated on the clinical validation aspect.
The methodology chosen to clinically validate the system was for a representative of the ICT to formulate a series of natural language queries; each one of these was annotated with the specific queried variable in the network and its value together with the known values of other variables.
The ICT's delegate included the expected reply to the queries together with its probability, based on the latest medical literature and their personal expertise.
So, each query asked for the probability truth value of the following:
\begin{align}
	(var_1 = a \wedge \ldots \wedge var_n = z ) \Rightarrow var_k = k \quad var_1 \ldots var_n \in V, var_k \in V \smallsetminus \{ var_1 \ldots var_n \}
\end{align}
This would translate into natural language into:
\enquote{If $var_1$ is $a$ and $\ldots$ and $var_n$ is $z$, how probable is it that $var_k$ $k$?}.
The complete series of eight questions, together with the expected values, is shown in Tab. \ref{tab:clinicalvalidationquestions}.
It is quite evident that all these validation questions are instances of the Bayesian Network Updating problem, in particular they are Conditional Probability Queries (as defined in \ref{subsec:bnupdating}).

\todo{inserire tabella qui o in Results}

\subsection{Explainability validation} \label{subsec:explainability-validation}
Having validated the system in terms of its adherence to clinical literature, it could then also meaningfully be validated from an explainability point of view.
The main question to be addressed is its capacity to relate to the expert user.
Is the system able to engender the user's trust?
In doing so, is she able to extract more knowledge from existing data when using the system than not?
Especially in cases where there may be a dearth of data, can the expert maximise the benefit from the available information?
Does the user subjectively feel that the system may positively impact her work?
These are all hard questions to answer, as there is a very high degree of subjectivity involved.
Thus to attempt to answer them, the chosen methods were borrowed from the social sciences.

In an earlier stage, the experts were introduced to the system in prototype form and instructed on the possible use cases it offered.
Then, they were asked to keep track of times they felt they could have used the system, had it been available, during their day-to-day work and to report back to me with the query they would have submitted to the system.
If the query were in the realm of possibilities offered by my program, I would submit the outputs back to the users together with the following questionnaire:
\todo[inline]{definire questionario da mandare con le domande}
This process would both give feedback on the performance of the system and also inform its design.

In a later phase, having finalised the system's design, I provided it to the experts at the ICT for their use in their daily work.
After \textbf{xx time} I followed up with them with an interview based on the following format:
\todo[inline]{definire formato e metriche intervista}
This enabled me to quantify the performance of the system, as perceived by its users in a real setting, over an extended period of time.

### 2019-07-05 methodology/instruments.tex
Sezione sull'algoritmo Pseudo-MPE

The full procedure can be found in \cite{koller2007dseparation}; my implementation shown in \ref{alg:d-separation} follows this pseudocode very closely but the procedure \texttt{d-separated}, instead of finding all nodes $R$ that are d-connected to the input $X$, tests if a given target $Y$ is d-separated from $X$ or not.
This gives some extra flexibility in how the function can be used.
To find the set $S$ of all nodes d-separated from $X$ I simply iterate the \texttt{d-separated} test over all nodes $V$ in the graph representing the BN.

\begin{algorithm}[htp!]
	\caption{d-separation algorithm}
	\label{alg:d-separation}
	\begin{algorithmic}
		\STATE $separated\_list = \emptyset$
		\STATE {\bf input:} source $X$, evidence $E$, nodes $V$
		\FOR{target $Y \in V$} 
		\STATE append $d-separated(X, Y, E)$ to $separated\_list$ \COMMENT{will return true or false}
		\ENDFOR
		\STATE {\bf output:} $separated\_list$
	\end{algorithmic}
\end{algorithm}

>>>>

The full procedure can be found in \cite{koller2007dseparation}; my implementation shown in Alg. \ref{alg:d-separation} follows this pseudocode very closely but the procedure \texttt{d-separated}, instead of finding all nodes $R$ that are d-connected to the input $X$, tests if a given target $Y$ is d-separated from $X$ or not.
This gives some extra flexibility in how the function can be used.
To find the set $S$ of all nodes d-separated from $X$ I simply iterate the \texttt{d-separated} test over all nodes $V$ in the graph representing the BN.

\begin{algorithm}[htp!]
	\caption{d-separation algorithm}
	\label{alg:d-separation}
	\begin{algorithmic}
		\STATE {\bf input:} source $X$, evidence $E$, nodes $V$
		\STATE $separated\_list = \emptyset$
		\FOR{target $Y \in V$} 
		\STATE append $d-separated(X, Y, E)$ to $separated\_list$ \COMMENT{will return true or false}
		\ENDFOR
		\STATE {\bf output:} $separated\_list$
	\end{algorithmic}
\end{algorithm}

### 2019-07-05 methodology/novel-contributions.tex
Sezione sull'algoritmo Pseudo-MPE

\subsection{Theory}

@@

\subsubsection{\enquote{Pseudo-MPE}} \label{subsubsec:pseudo-mpe}
\todo{dove critico il fatto che non calcolano veramente l'MPE come dicevano? qui o in cap.4?}
\todo{eseguibile in modalita' esaustiva, d-separata o thresholdata}

@@

\subsubsection{Plot of network} \label{subsubsec:plot-model}

>>>>

\subsection{Theory} \label{subsec:theory}
\todo{la discussione con l'esperto ci da' belief revision dei dati}
\todo{counterfactual explanations}

\subsubsection{Selection based on entropy}
\todo{utilizzo entropia}

@@

\subsubsection{\enquote{Pseudo-MPE}} \label{subsubsec:pseudo-mpe}
\todo{dove critico il fatto che non calcolano veramente l'MPE come dicevano? qui o in cap.4 on cap.2?}
The so-called \enquote{Pseudo-MPE} algorithm is inherently wrapped up with the concept of \textit{dialogue} and is central to the explanatory powers of the system being developed in this thesis.
The algorithm was developed as a way of implementing the \enquote{MPE branch} of the \enquote{Argumentative Probability Tree} hypothesised by \cite{Butz2018}.
It was termed \enquote{Pseudo-MPE} because there are no guarantees of it returning the MPE solution (see Subsec. \ref{subsec:bnupdating}), as noted by \cite{koller2007introduction} in their definition of the MAP problem.

At a lower level of detail, the algorithm may be broken into:
\begin{itemize}
	\item a dialogical part, that interfaces with the expert user through the use of natural language, menus and visualisations
	\item the part responsible for constructing the \enquote{MPE branch}
\end{itemize}
The former process was informed and shaped by the results obtained by the methods described in Subsec. \ref{subsec:explainability-validation} and, as such, presented substantial elements of novelty.

The latter process is, at its core, a greedy procedure that aims at selecting the \enquote{best} next $(state, value)$ tuple at each step, based on some measure of optimality and on the variables already in the evidence set.
In the actual implemented system the two parts are intertwined, given their close inter-dependence.

The \texttt{dialogue} procedure starts by asking the user to select a subset of variables and their relative values to add as initial evidence.
This initial evidence is used to radicate the MPE Branch.
It should be noted than in the description given by \cite{Butz2018}, the Argumentative Probability Tree is a real tree as each node is guaranteed to have at most one parent.
My application, on the other hand, constructs an \textit{Argumentative Probability Polytree} (see Subsec. \ref{subsec:graph-theory}) because, as will better be described in Chap. \ref{chap:results}, it was seen early on that the users much preferred to be able to start from a set of initial evidences and not be limited to a single one.
The algorithm then proceeds to call the \texttt{next\_most\_probable\_states} subroutine that is tasked with returning an ordered list of $(state,value)$ pairs.
It does this by calculating the posterior distribution given evidence of all the states not already in the evidence, then calculating the efficiency (see \ref{subsec:information-theory}) and the maximally probable symbol of each state's distribution and finally returning the $(state,value)$ tuples ordered according to their normalised entropy (most efficient/least entropic at the head).
The $(state,value)$ pair at the head of the list is proposed to the user who has the faculty to accept the system's evaluation or refuse it.
If the user accepts, the state is added to the evidence set and to the MPE Branch under construction.
Thus, the evidence set's cardinality increases by one each time a user accepts a proposal.
The updated evidence will be used to calculate the new list of $(state,value)$ pairs at the following round.
If the expert chooses to refuse, then she is iteratively presented with the remaining $(state,value)$ items, in order of decreasing efficiency. 
Once she accepts one of the explanations given by the system, the \texttt{generate\_alternative\_branch} subroutine is called to automatically generate a maximally probable MPE Branch, radicated in the newest $(state,value)$ node of the MPE Branch.
The proposal loop for alternative states runs until there are increasingly less probable elements in the list and exits with a partial solution if the user refuses all of them at a given step.
Thus, the Pseudo-MPE solution is constructed only if the user runs through all variables proposed, accepting each one at least once.

Three slightly different operational modes of the algorithm were implemented.
This was done for research purposes, in order to understand which of the three, if any or if a combination of their distinctive features, the expert users would find the most useful from a usability, comprensibility and explainability standpoint:
\begin{itemize}
  \item exhaustive: In the basic dialogue type, the set of variables under consideration monotonically decreases by one every time the user accepts one of the system's proposals and the dialogue terminates only when the user has accepted all variables at least once or refused all proposals at a given step.
	In the first case the user will have the Pseudo-MPE solution while in the second she will be left with a partial assignment to some of the variables not present in the initial evidence.
	The pseudocode is shown in Alg. \ref{alg:pseudo-mpe-exhaustive}.
  \item d-separated: In the second variant, the set of variables under considered at each step is dynamic and depends on the separation properties of the underlying Bayesian Network's DAG and the evidence set constructed by the user's choices.
  	Differently from the first type of dialogue, an additional \texttt{evidence\_d\_separation} subroutine is called before \texttt{next\_most\_probable\_states} to calculate the set of variables that are d-separated from the evidence set, up to that step of the dialogue.
  	\texttt{next\_most\_probable\_states} is then executed but the variables that the previous function found to be separated from the evidence, are removed from the returned list.
  	This way, variables that can have no effect given the current evidence are not proposed.
  	As the d-separation operation is not monotonic, adding new nodes to the evidence set can both increment or diminish the number of nodes that will be proposed at each step.
  	The user is shown an updated view of the independencies of the graph at each step; an example of such an output is shown in Fig. \ref{fig:pseudo-mpe-independencies}
  	The pseudocode is shown in Alg. \ref{alg:pseudo-mpe-independencies}.
  \item thresholded: The final variant of the algorithm prunes the set of variables using a different strategy from the previously presented one.
  	In this case, the $(state,value)$ pairs in the list returned by \texttt{next\_most\_probable\_states} are dropped or automatically accepted based on their normalised entropy score.
  	Pairs whose entropy is above a user-defined threshold are removed and not proposed to the user while, conversely, tuples with a low enough score are automatically accepted and added to the MPE Graph and the evidence set.
  	This thresholding strategy based on the efficiency of the variables is paired with one where the threshold considers the number of times that the expert has refused a given state. Tuples can be proposed multiple times, with an ever lower probability, if the user has refused them previously; in this scheme a $(state,value)$ pair can only be proposed a maximum number of times.
\end{itemize}

\begin{algorithm}[htp!]
	\caption{Exhaustive pseudo-MPE algorithm}
	\label{alg:pseudo-mpe-exhaustive}
	\begin{algorithmic}
		\STATE $evidence = $ user selected $(state,value)$ tuples
		\STATE $MPE_tree = $ root MPE Tree in $evidence$
		\WHILE{True} 
			\STATE $mpe\_states$ = \texttt{next\_most\_probable\_states()}
			\IF{$mpe\_states$ is not empty}
				\STATE $next\_state$ = head of $mpe\_states$ 
				\STATE propose $next\_state$ to user \COMMENT{the least entropic state}
				\IF{the user refuses $next\_state$}
					\FOR{$alternative\_state$ in $mpe\_states \smallsetminus next\_state$}
						\STATE propose $alternative\_state$ to user \COMMENT{the next least entropic states}
						\IF{the user accepts $alternative\_state$}
							\STATE call \texttt{generate\_alternative\_branch()} on $MPE\_tree$ 
							\STATE add $alternative\_state$ to $MPE\_tree$
							\STATE $evidence = evidence \cup alternative\_state$
						\ELSE
							\STATE continue
						\ENDIF
					\ENDFOR
				\ELSE
					\STATE add $next\_state$ to $MPE\_tree$
					\STATE $evidence = evidence \cup next\_state$
				\ENDIF
			\ELSE 
				\STATE return
			\ENDIF
		\ENDWHILE
	\end{algorithmic}
\end{algorithm}

\begin{algorithm}[htp!]
	\caption{Independencies pseudo-MPE algorithm}
	\label{alg:pseudo-mpe-independencies}
	\begin{algorithmic}
		\STATE $evidence = $ user selected $(state,value)$ tuples
		\STATE $MPE_tree = $ root MPE Tree in $evidence$
		\WHILE{True} 
			\STATE $separated = $ \texttt{evidence\_d\_separation()} \COMMENT{based on evidence of previous step}
			\STATE $mpe\_states$ = \texttt{next\_most\_probable\_states()}
			\STATE $mpe\_states = mpe\_states \smallsetminus separated$ 
			\IF{$mpe\_states$ is not empty}
				\STATE $next\_state$ = head of $mpe\_states$ 
				\STATE propose $next\_state$ to user \COMMENT{the least entropic state}
				\IF{the user refuses $next\_state$}
					\FOR{$alternative\_state$ in $mpe\_states \smallsetminus next\_state$}
						\STATE propose $alternative\_state$ to user \COMMENT{the next least entropic states}
						\IF{the user accepts $alternative\_state$}
							\STATE call \texttt{generate\_alternative\_branch()} on $MPE\_tree$ 
							\STATE add $alternative\_state$ to $MPE\_tree$
							\STATE $evidence = evidence \cup alternative\_state$
						\ELSE
							\STATE continue
						\ENDIF
					\ENDFOR
				\ELSE
					\STATE add $next\_state$ to $MPE\_tree$
					\STATE $evidence = evidence \cup next\_state$
				\ENDIF
			\ELSE 
				\STATE return
			\ENDIF
		\ENDWHILE
	\end{algorithmic}
\end{algorithm}

\begin{figure}[htbp]
\centerline{\includegraphics[width=\columnwidth]{methodology/images/example-d-separation-mpe_1}}
\caption{Example output during the first round of the d-separation-aware variant of \texttt{dialogue}.
	The variable \enquote{mut17q21} is the initial evidence.}
\label{fig:pseudo-mpe-independencies_1}
\end{figure}

\begin{figure}[htbp]
\centerline{\includegraphics[width=\columnwidth]{methodology/images/example-d-separation-mpe_2}}
\caption{Example output during the second round of the d-separation-aware variant of \texttt{dialogue}.
	\enquote{morfologia} is added to the evidence set and this makes a large part of the network redundant.}
\label{fig:pseudo-mpe-independencies_2}
\end{figure}

\todo{implementare e finire}
\begin{algorithm}[htp!]
	\caption{Thresholded pseudo-MPE algorithm}
	\label{alg:pseudo-mpe-thresholded}
	\begin{algorithmic}
		\STATE $separated\_list = \emptyset$
		\STATE {\bf input:} source $X$, evidence $E$, nodes $V$
		\FOR{target $Y \in V$} 
		\STATE append $d-separated(X, Y, E)$ to $separated\_list$ \COMMENT{will return true or false}
		\ENDFOR
		\STATE {\bf output:} $separated\_list$
	\end{algorithmic}
\end{algorithm}

@@

\subsubsection{Network Plotting} \label{subsubsec:plot-model}
